---
title: "20230714_If_else_RClub"
author: "Samuel Gurr"
date: "2023-06-22"
output: html_document
---

# If / Else Workshop

OBJECTIVE:

-   use conditional statements to manipulate large datasets, converting values based on binary criteria

    -   a custom tutorial is provided to become familiar with if else statments!

    -   several questions for group work

-   in summary, if else statments represent a simple yet powerful tool that demistifies the 'ones and zeros' of coding. If you can logically reach your desired outcome from a series of boolean values (binary as true and false), you can do anything!

### Load Libraries

-   NOTE: as opposed to running the whole chunk, position your cursor on each line and use Cnrtl+Enter to run line by line. if you get a warning message that you do not have the package, use install.package() command

```{r startup, include=FALSE}

```

### Using a mock dataset 'ToothGrowth'

ToothGrowth data set contains the result from an experiment studying the effect of vitamin C on tooth growth in 60 Guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

```{r load data}
data() # take a look! these are all mock datasets in base R! 

head(ToothGrowth) # first size lines of the dataset we will use

colnames(ToothGrowth) # view the column names len supp and dose

ToothGrowth$supp  # string categorical variable  - these are vitamin C 'VC' and orange juice 'OJ'
ToothGrowth$len   # continues variable  - the length of Guinea pig teeth
ToothGrowth$dose  # numeric categorical variable   - the level of supplement 
```

### Before getting started..

Lets assign DF_tooth as our working version of ToothGrowth in our exercises

```{r}
DF_tooth <- ToothGrowth
```

### Conditional statements

About: if else statements following a boolean logic, meaning a series of binary true and false outcomes. They are are coded in R as the following:

**if** ( statement here ) {

outcome if true

} **else** { outcome if false }

```{r}

x = 1.1 # change me and run again below! 

if (x < 1) { # if x is < than 1...
  
  print('x is less than 1') # print this message if TRUE 
  
} else { # else means here than x is actually greater than one
  
  print('x is greater than 1') # print this message if FALSE 
  
}

```

```{r}


if (DF_tooth$len > 20) { # if x is < than 1...
  
  DF_tooth$note <- "Wowza, huge teeth!"  # print this message if TRUE
  
} else { # else means here than x is actually greater than one
  
  DF_tooth$note <- "regular unsuprisingly-sized teeth.."  # print this message if FALSE 
  
}


DF_tooth$note  <- ifelse(DF_tooth$len > 20, "Wowza, huge chompers!",  "regular unsuprisingly-sized teeth.." )
View(DF_tooth)
```

```{r}


DF_tooth$len <- ifelse(df$team == 'A', 'great', 'bad')


```

```{r}

```

(1) Turns out they feed the Guinea pigs orange juice with extra pulp! Therfore the tooth measurements were overestimated by 10%

```{r}

```
