---
title: "20230714_If_else_RClub"
author: "Samuel Gurr"
date: "2023-07-14"
output: html_document
---

# If / Else Workshop

## Click 'Visual' at the top of this window for the formatted experience 

'if else' statements are a simple yet powerful tool that can demystify the 'ones and zeros' of coding. If you can logically reach your desired outcome from a series of true and false statement, you can do anything!

**OBJECTIVE:**

-   use conditional statements to control large data sets, converting values based on conditional criteria

    -   a custom tutorial is provided to become familiar with if else statements!

    -   a few questions for group work

-   familiarize ourselves with common logical operators

    -   `!` = NOT

    -   `&` = AND (element wise)

    -   `|` = OR (element wise)

    -   note: element wise meaning the operators performs the if you double up for && and //

## Intro to boolean variables

-   boolean (computing definition) - a binary variable or outcome as 'true' or 'false'

**Conditional statements -** if else statements following a boolean logic, meaning a series of binary true and false outcomes. They are are coded in R as the following:

**if** ( statement here ) {

outcome if true

} **else** { outcome if false }

**Let's give it a go shall we!**

-   run line by line to understand the boolean logic (true and false)

-   change x to receive the opposite outcome

```{r intro to boolean, include=TRUE}

x = 1.1 # change me and run again below! 

if (x < 1) { # if x is < than 1...
  
  print('x is less than 1') # print this message if TRUE 
  
} else { # else means here than x is  greater than one
  
  print('x is greater than 1') # print this message if FALSE 
  
}

```

## Operating on if else with Operators

Problem: We have a conditional statement based on a threshold that is a likely outcome.

X=1 is the metaphorical wrench that will break our if else machine!

-   run the code below

```{r throw the wrench, some more!, include=TRUE}

x = 1 # the wrench

# our machine
if (x < 1) { # if x is < than 1…    
  
  print('x is less than 1') # print this message if TRUE    
  
} else { # else means here than x is  greater than one    
  
    print('x is greater than 1') # print this message if FALSE  
  
  }

```

-   "x is greater than 1" but... but... this is FALSE! ("oh no!")

    -   the `if` statement asks if x is less than 1 -- it was not, it simply ran the remaining `else` output

### Use operators to save us from doom

option 1: using `|` for OR statement

```{r help yourself with OR, some more!, include=TRUE}

x = 1 # the value that throws the metaforical wrench on our code 

if (x < 1 | x == 1) { # add  the OR statement!
  
  print('x is less than or equal to 1') # change your output to inform yourself
  
} else { # else means here than x is  greater than one
  
  print('x is greater than 1') # print this message if FALSE 
}

```

```{r help yourself with AND, some more!, include=TRUE}

x = 1 

if (x < 1 & x != 1) { # add & for x not equal to 1
  
  print('x is less than 1') # added more information for this output
  
} else { # else means here than x is  greater than one
  
  print('x is greater than or equal to 1') # print this message if FALSE 
}
```

option 2: using `&` or AND statement

```{r help yourself with AND, some more!, include=TRUE}

x = 1 

if (x < 1 & x != 1) { # add & for x not equal to 1
  
  print('x is less than 1') # added more information for this output
  
} else { # else means here than x is  greater than one
  
  print('x is greater than or equal to 1') # print this message if FALSE 
}
```

#### What if you want to add another statement? use `else if`

`else if` - if your first `if` statement is false, `else if` provides another opportunity before defaulting to the `else` output

-   note you can have as many `else if` statements that you's like - just remember to keep organized and logical

option 3: add `else if`

```{r else... if..?, some more!, include=TRUE}
x = 1   

if (x < 1 ) { # if x is lass than 1      
  
  print('x is less than 1') # output if TRUE 

} else if (x == 1) {
  
  print('x is equal to 1, wrench removed - your machine lives')
  
} else { # else means here than x is  greater than one     
    
  print('x is greater than or equal to 1') # print this message if FALSE  
  
  }
```

### Activity: Sunday scaries (ooooo!)

#### *Imagine* it is mid-July and the weather has been beautiful and warm. You had an amazing weekend and you don't want it to end.

#### "...oh ..hmm that *muuust* be next week ..oh no!"

#### A jolt of shivers runs up your spine, it is after 5 pm on Sunday already! You just remembered you are leading a meeting tomorrow morning!

-   Objective: edit the time and your preparedness to receive the output `blame it on the kids and stay home tomorrow`
-   Understand the `if`, `else, and else if` conditional statement with operators to solve!
-   a chunk is created below, fill the variables and learn your destiny

```{r sunday scaries, include=TRUE}
#  install.packages('lubridate') # if you do not have it already
library(lubridate) # converts the hour and minutes you input into an integer, 


# assign your variables
day    =  # hmm.. what day is it?

time   =  as.numeric(hm('00:00')) # input military time, curently midnight 'hm' lubridate

ready  =   # add whether you are ready as 'yes' or 'no' o include  the quptes as a string


# run the conditional statement and learn your destiny 

if (day != 'Sunday' & ready == 'no') { # the pseudosunday scaries 
  
  print('Phew! Blissfully procrastinate more') 
  
} else if (day == 'Sunday' & ready != 'no' & time < as.numeric(hm('17:00'))) { # almost forgot that you are awesome
  
  print('Oh yeah - you prepared last week! You are a badass, enjoy your time in the sun.')
  
} else if (day == 'Sunday' & ready != 'no' & time > as.numeric(hm('17:00'))) {
  
  print('Popcorn time')
  
} else if (day == 'Sunday' & ready == 'no' & time < as.numeric(hm('17:00'))) {
  
  print('Palms sweaty, eyes twitching - become awkwardly quiet and stiff. You scramble to finish your wide-eyed goodbyes to your friends as you inch closer to the exit. The computer calls our to you, finish the task if your life depends on it.')
  
} else { 
  
  print('Just blame it on the kids and stay home tomorrow') # print this message if FALSE 
}
```

### 
